---
title: "08-deseq2.Rmd"
output: html_document
date: "2023-12-14"
---
Rmd to use `DESeq2` to get differentially expressed genes lists from summmer 2022 samples. Experiment was comparing the response to exposure of SSWD between small (juvenile) and large (adult) _Pycnopodia helianthoides_ and compared to control stars. 

Sample info: 
https://github.com/grace-ac/project-pycno-sizeclass-2022/blob/main/data/summer2022_metadata.csv 
Load packages:    
```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(data.table)
```

Read in count matrix created on Raven using: https://github.com/grace-ac/project-pycno-sizeclass-2022/blob/main/code/07-kallisto-summer2022-phelgenome.Rmd 

```{r}
countmatrix <- read.delim("../analyses/07-kallisto-phelgenome/kallisto_20231204.isoform.counts.matrix", header = TRUE, sep = '\t')
rownames(countmatrix) <- countmatrix$X
countmatrix <- countmatrix[,-1]
head(countmatrix)
```

Round integers up to hole numbers for further analysis:
```{r}
countmatrix <- round(countmatrix, 0)
head(countmatrix)
```

# Compare: Exposed Large Adult _Pycnopodia helianthoides_ (n=4) to Control Large Adult _Pycnopodia helianthoides_ (n=4)

Samples:
Exposed: PSC.0190; PSC.0206; PSC.0217; PSC.0231     
Control: PSC.0186; PSC.0202; PSC.0203; PSC.0209    

Pull out those samples:
```{r}
#levlc is large exposed vs large control
levlc <- select(countmatrix, "PSC.0190", "PSC.0206", "PSC.0217", "PSC.0231", "PSC.0186", "PSC.0202", "PSC.0203", "PSC.0209")
head(levlc)
```

Get DEGs based on exposure to SSWD / disease sign (all exposed adults had at least 1 dropped arm and some had twisting)

```{r}
colData <- data.frame(condition=factor(c("exposed", "exposed", "exposed", "exposed", "control", "control", "control", "control")), 
                             type=factor(rep("paired-end", 8)))
rownames(colData) <- colnames(levlc)
dds <- DESeqDataSetFromMatrix(countData = levlc,
                                     colData = colData, 
                                     design = ~ condition)
```

```{r}
dds <- DESeq(dds)
res <- results(dds)
res <- res[order(rownames(res)), ]
```

```{r}
head(res)
```

```{r}
# Count number of hits with adjusted p-value less then 0.05
dim(res[!is.na(res$padj) & res$padj <= 0.05, ])
```
[1] 4617    6

```{r}
adult_exposure_fig <- res
# The main plot
plot(adult_exposure_fig$baseMean, adult_exposure_fig$log2FoldChange, pch=20, cex=0.45, ylim=c(-15, 15), log="x", col="darkgray",
     main="SSWD Exposure Status  (pval </= 0.05)",
     xlab="mean of normalized counts",
     ylab="Log2 Fold Change")
# Getting the significant points and plotting them again so they're a different color
adult_exposure_fig.sig <- res[!is.na(res$padj) & res$padj <= 0.05, ]
points(adult_exposure_fig.sig$baseMean, adult_exposure_fig.sig$log2FoldChange, pch=20, cex=0.45, col="red")
# 2 FC lines
abline(h=c(-1,1), col="blue")
```

# Compare: Exposed Small Juvenile _Pycnopodia helianthoides_ (n=4) to Control Small Juvenile _Pycnopodia helianthoides_ (n=4)









