---
title: "09-PCAplots.Rmd"
output: html_document
date: "2023-12-19"
---
Rmd to make some PCA plots of summer 2022 RNAseq data to help figure out what `DESeq2` comparisons to make. 


New version of RStudio downloaded Dec 19, 2023 
```{r}
sessionInfo()
```

# Start with PCA plot of all 32 libraries  
```{r}
#commented out chunk 20231219 bc have run already
#if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("DESeq2")

```

Load packages:
```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(dplyr)
```

# Read in count matrix from comparing the 2022 _Pycnopodia helianthoides_ coelomocyte RNAseq libraries with the 2015 _Pycnopodia helianthoides_ gene list of the published genome:  
```{r}
countmatrix <- read.delim("../analyses/07-kallisto-phelgenome/kallisto_20231204.isoform.counts.matrix", header = TRUE, sep = '\t')
rownames(countmatrix) <- countmatrix$X
countmatrix <- countmatrix[,-1]
head(countmatrix)
```
26581 rows, 32 columns 

Round integers up to whole numbers for analyses:    
```{r}
countmatrix <- round(countmatrix, 0)
head(countmatrix)
```

write out matrix of rounded up integers: 
Wrote out 2023-12-20 1800pm
```{r}
#write.table(countmatrix, "../analyses/07-kallisto-phelgenome/32libraries_counts_rounded.tsv", sep = "\t", row.names = TRUE, quote = FALSE)
```



Make a data frame for the comparison: 
```{r}
colData <- data.frame(condition=factor(c("day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0S", "day0S", "day0S", "day0S", "day8LA",  "day8LA", "day9LA2", "day10LA", "day11SH", "day11SH", "day11SA", "day11SA", "day11LA", "day12SA", "day12LH", "day12LH", "day12LA2", "day13LH", "day13LA9", "day14SH", "day14SA", "day15SH", "day15LA", "day15SD")),
                      type=factor(rep("paired-end",32)))
rownames(colData) <- colnames(countmatrix)
dds <- DESeqDataSetFromMatrix(countData = countmatrix,
                              colData = colData,
                              design = ~ condition)
```
```{r}
dds <- DESeq(dds)
res <- results(dds)
res <- res[order(rownames(res)), ]
```

From [Bioconductor `DESeq2` Vignette](http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html):     
```{r}
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)
```


```{r}
plot <- plotPCA(vsd, intgroup=c("condition", "type"))
nudge <- position_nudge(y = 4)
plot + geom_text(aes(label = name), position = nudge)
```

That's kind of messy and confusing... 

# Adonis PCA to see if individual treatment and days since exposure interacts with disease sign 

```{r}
#comment out code 20231219
#install.packages("vegan")
```

```{r}
library(vegan)

help(`adonis2`)
```

So based on notes I have from December 4, 2023 Pycno meeting with Alyssa and Melanie, I'm going to: 

All first arm drop ones
Do they all look the same or different

How does individual treatment (size/age) and date interact with disease sign 
https://rdrr.io/rforge/vegan/man/adonis.html 
Response variable treatment by sign of disease by day
Interactive 
If that was significant, then that would say there’s a difference in sign of disease 
If significance, need to to pay attention to date since the exposure rather than the sign 
 
| coelom_ID | star_ID | star_size | treatment_grp | experiment_day | disease_sign                        |
|-----------|---------|-----------|---------------|----------------|-------------------------------------|
| PSC 0149  |      28 | large     | exposed       |              8 | FIRST arm dropped, 1 arm twisted    |
| PSC 0150  |       6 | large     | exposed       |              8 | FIRST arm dropped, 4 arms twisted   |
| PSC 0174  |      30 | large     | exposed       |             10 | FIRST arm dropped, 6 arms twisted   |
| PSC 0187  |      57 | small     | exposed       |             11 | 1 arm dropped, twisting, stretching |
| PSC 0188  |      38 | small     | exposed       |             11 | 1 arm dropped, stretching           |
| PSC 0190  |      10 | large     | exposed       |             11 | FIRST arm dropped                   |
| PSC 0228  |      61 | small     | exposed       |             14 | 1 arm dropped                       |
| PSC 0231  |      13 | large     | exposed       |             15 | FIRST arm dropped, 5 arms twisted   |


Make a subset countmatrix of just those libraries: 
```{r}
adcmatrix <- select(countmatrix, "PSC.0149", "PSC.0150", "PSC.0174", "PSC.0187", "PSC.0188", "PSC.0190", "PSC.0228", "PSC.0231")
head(adcmatrix)
```
26581 rows, 8 columns

write out table
```{r}
write.table(adcmatrix, "../analyses/07-kallisto-phelgenome/firstarmdrop_counts_rounded.tsv", sep = "\t", row.names = TRUE, quote = FALSE)
```


### Make a regular PCA plot first. 

Make a data frame for the comparison: 
```{r}
colData <- data.frame(condition=factor(c("day8L","day8L","day10L","day11S","day11S","day11L","day14S","day15L")),
                      type=factor(rep("paired-end",8)))
rownames(colData) <- colnames(adcmatrix)
dds <- DESeqDataSetFromMatrix(countData = adcmatrix,
                              colData = colData,
                              design = ~ condition)
```
```{r}
dds <- DESeq(dds)
res <- results(dds)
res <- res[order(rownames(res)), ]
```

From [Bioconductor `DESeq2` Vignette](http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html):     
```{r}
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)
```
```{r}
plotad <- plotPCA(vsd, intgroup=c("condition", "type"))
nudge <- position_nudge(y = 4)
plotad + geom_text(aes(label = name), position = nudge)
```

## Try using `adonis2` to make plot comparing RNAseq data - see if age/size of pycno (large or small) and days since exposure impacts the disease sign (arm drop) 

I think the data need to be transposed... the columns should be the genes, and the rows should be the 8 library names

make the gene list rownames in to a column: 
```{r}
library(tibble)
dfadcmatrix <- tibble::rownames_to_column(adcmatrix, "gene")
head(dfadcmatrix)
```

make first column a no-named column:
https://stackoverflow.com/questions/53493342/remove-the-first-column-name-in-a-data-frame-from-fread-in-r 
```{r}
#names(dfadcmatrix)[1] <- ""
#head(dfadcmatrix)
```


```{r}
#tadcmatrix <- transpose(dfadcmatrix)
#head(tadcmatrix)
```

then make the first row into the column names: 
```{r}
#names(tadcmatrix) <- tadcmatrix[1,]
```

```{r}
#tadcmatrix <- tadcmatrix[-1,]
#head(tadcmatrix)
```

then make the row names back into the library names: 
PCA.0149; PCA.0150; PCA.0174; PCA.0187; PCA.0188; PCA.0190; PCA.0228; PCA.0231

```{r}
#rownames(tadcmatrix) <- c("PCA.0149", "PCA.0150", "PCA.0174", "PCA.0187", "PCA.0188", "PCA.0190", "PCA.0228", "PCA.0231")
#head(tadcmatrix)
```

name the first column "Library"
```{r}
#library(tibble)
#tadcmatrix <- tibble::rownames_to_column(tadcmatrix, "Library")
#head(tadcmatrix)
```


```{r}
#adonis2(adcmatrix ~ )
```


```{r}
#armdrop.pca <- prcomp(adcmatrix, 
                   center = TRUE, 
                   scale. = TRUE) 
  
# summary of the  
# prcomp object 
#summary(armdrop.pca)
```
# structure of the pca object 
```{r}
#str(armdrop.pca) 
```


# loading library 
```{r}
#library(ggfortify) 
#?autoplot
#armdrop.pca.plot <- autoplot(armdrop.pca, 
                          #data = adcmatrix,
                          #colour = 'Library') 
  
#armdrop.pca.plot
```

# Try using `pcaExplorer` from BioConductor
https://bioconductor.org/packages/release/bioc/vignettes/pcaExplorer/inst/doc/pcaExplorer.html 

https://bioconductor.org/packages/3.18/bioc/vignettes/pcaExplorer/inst/doc/upandrunning.html 
Installed 2023-12-20
```{r}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("pcaExplorer")
```

```{r}
#BiocManager::install("pcaExplorer", dependencies = TRUE)
```

installed 2023-12-20
```{r}
#install.packages("markdown")
```
installed 2023-12-20
```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("BiocStyle")
```

```{r}
library(BiocStyle)
```



```{r}
library("pcaExplorer")
pcaExplorer()
```
The count matrix file is analyses/07-kallisto-phelgenome/firstarmdrop_counts_rounded.tsv

The sample metadata matrix file is data/firstarmdrop_coldata.tsv 

Saved the pdf images to analyses/09-PCAplots



## Try using `pcaExplorer` with all 32 libraries: 
```{r}
library("pcaExplorer")
pcaExplorer()
```


input files: 
count matrix file: analyses/07-kallisto-phelgenome/32libraries_counts_rounded.tsv
metadata matrix: data/summer2022_coldata.tsv

save pdf images and report to: analyses/09-PCAplots

## play with the airway data from `pcaExplorer` to see if i can figure out how to format my data to make a multifactor comparison

```{r}
BiocManager::install("airway")
```

```{r}
library("pcaExplorer")
pcaExplorer()
```

comment out 2023-12-21 because has been installed
```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("org.Hs.eg.db")
```



