---
title: "09-PCAplots.Rmd"
output: html_document
date: "2023-12-19"
---
Rmd to make some PCA plots of summer 2022 RNAseq data to help figure out what `DESeq2` comparisons to make. 


New version of RStudio downloaded Dec 19, 2023 
```{r}
sessionInfo()
```

# Start with PCA plot of all 32 libraries  
```{r}
#commented out chunk 20231219 bc have run already
#if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("DESeq2")

```

Load packages:
```{r}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
library(data.table)
library(ggplot2)
library(dplyr)
```

# Read in count matrix from comparing the 2022 _Pycnopodia helianthoides_ coelomocyte RNAseq libraries with the 2015 _Pycnopodia helianthoides_ gene list of the published genome:  
```{r}
countmatrix <- read.delim("../analyses/07-kallisto-phelgenome/kallisto_20231204.isoform.counts.matrix", header = TRUE, sep = '\t')
rownames(countmatrix) <- countmatrix$X
countmatrix <- countmatrix[,-1]
head(countmatrix)
```
26581 rows, 32 columns 

Round integers up to whole numbers for analyses:    
```{r}
countmatrix <- round(countmatrix, 0)
head(countmatrix)
```

Make a data frame for the comparison: 
```{r}
colData <- data.frame(condition=factor(c("day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0L", "day0S", "day0S", "day0S", "day0S", "day8LA",  "day8LA", "day9LA2", "day10LA", "day11SH", "day11SH", "day11SA", "day11SA", "day11LA", "day12SA", "day12LH", "day12LH", "day12LA2", "day13LH", "day13LA9", "day14SH", "day14SA", "day15SH", "day15LA", "day15SD")),
                      type=factor(rep("paired-end",32)))
rownames(colData) <- colnames(countmatrix)
dds <- DESeqDataSetFromMatrix(countData = countmatrix,
                              colData = colData,
                              design = ~ condition)
```
```{r}
dds <- DESeq(dds)
res <- results(dds)
res <- res[order(rownames(res)), ]
```

From [Bioconductor `DESeq2` Vignette](http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html):     
```{r}
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)
```


```{r}
plot <- plotPCA(vsd, intgroup=c("condition", "type"))
nudge <- position_nudge(y = 4)
plot + geom_text(aes(label = name), position = nudge)
```

That's kind of messy and confusing... 

# Adonis PCA to see if individual treatment and days since exposure interacts with disease sign 

```{r}
#comment out code 20231219
#install.packages("vegan")
```

```{r}
library(vegan)

help(`adonis2`)
```

So based on notes I have from December 4, 2023 Pycno meeting with Alyssa and Melanie, I'm going to: 

All first arm drop ones
Do they all look the same or different

How does individual treatment (size/age) and date interact with disease sign 
https://rdrr.io/rforge/vegan/man/adonis.html 
Response variable treatment by sign of disease by day
Interactive 
If that was significant, then that would say there’s a difference in sign of disease 
If significance, need to to pay attention to date since the exposure rather than the sign 
 
| coelom_ID | star_ID | star_size | treatment_grp | experiment_day | disease_sign                        |
|-----------|---------|-----------|---------------|----------------|-------------------------------------|
| PSC 0149  |      28 | large     | exposed       |              8 | FIRST arm dropped, 1 arm twisted    |
| PSC 0150  |       6 | large     | exposed       |              8 | FIRST arm dropped, 4 arms twisted   |
| PSC 0174  |      30 | large     | exposed       |             10 | FIRST arm dropped, 6 arms twisted   |
| PSC 0187  |      57 | small     | exposed       |             11 | 1 arm dropped, twisting, stretching |
| PSC 0188  |      38 | small     | exposed       |             11 | 1 arm dropped, stretching           |
| PSC 0190  |      10 | large     | exposed       |             11 | FIRST arm dropped                   |
| PSC 0228  |      61 | small     | exposed       |             14 | 1 arm dropped                       |
| PSC 0231  |      13 | large     | exposed       |             15 | FIRST arm dropped, 5 arms twisted   |


Make a subset countmatrix of just those libraries: 
```{r}
adcmatrix <- select(countmatrix, "PSC.0149", "PSC.0150", "PSC.0174", "PSC.0187", "PSC.0188", "PSC.0190", "PSC.0228", "PSC.0231")
head(adcmatrix)
```
26581 rows, 8 columns

### Make a regular PCA plot first. 

Make a data frame for the comparison: 
```{r}
colData <- data.frame(condition=factor(c("day8L","day8L","day10L","day11S","day11S","day11L","day14S","day15L")),
                      type=factor(rep("paired-end",8)))
rownames(colData) <- colnames(adcmatrix)
dds <- DESeqDataSetFromMatrix(countData = adcmatrix,
                              colData = colData,
                              design = ~ condition)
```
```{r}
dds <- DESeq(dds)
res <- results(dds)
res <- res[order(rownames(res)), ]
```

From [Bioconductor `DESeq2` Vignette](http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html):     
```{r}
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
head(assay(vsd), 3)
```
```{r}
plotPCA(vsd, intgroup=c("condition", "type"))
```

```{r}
#install.packages("ggfortify")
```

```{r}
library(ggfortify)

pca_res <- prcomp(adcmatrix, scale. = TRUE)

autoplot(pca_res, data = adcmatrix, colour = 'Treatment')
```





