---
title: "13-compareUpinArms-2022"
output: html_document
date: "2024-01-02"
---
Rmd to compare DEGlist from Up in Arms 2015 to Summer 2022 DEG list comparing 7control and 7 exposed arm drop stars. 

Load packages needed:
```{r}
library(tidyverse)
library(dplyr)
```

Read in Up in Arms DEGlist from https://github.com/sr320/eimd-sswd/blob/master/precompiled_wd/Phel_DEG_uniprot.tab. I downloaded the file and put in the data directory for this repository. 

```{r}
uiadeg <- read.delim("../data/Phel_DEG_uniprot_UpinArms.tab", sep = '\t', header = TRUE)
head(uiadeg)
```

Pull out just the first two columns:
```{r}
uiatuni <- select(uiadeg, "transcript_ID", "uniprotID")
head(uiatuni)
```

Read in the DEGlist with uniprot from the summer 2022 deseq2 comparison of 7controls vs 7 exposed arm drop stars:
```{r}
cvad <- read.delim("../analyses/08-deseq2/DEGlist_control_v_armdrop_7x7_genecounts_blast.tab")
head(cvad)
```

`join` this list with the uiatuni list based on the column "uniprotID"
```{r}
cvaduia <- left_join(cvad, uiatuni, by = "uniprotID")
head(cvaduia)
```

Remove rows with no input in uniprotID column:
newdf <- df[!is.na(df$name),]
```{r}
cvaduiashort <- cvaduia[!is.na(cvaduia$uniprotID),]
head(cvaduiashort)
```

Remove rows that don't have a Up in Arms transcript ID:
```{r}
cvaduiashorter <- cvaduiashort[!is.na(cvaduiashort$transcript_ID),]
head(cvaduiashorter)
```

It is now 1,727 rows long. 

Read this file into a analyses/13-compare-UpinArms:
```{r}
#write.table(cvaduiashorter, "../analyses/13-compare-UpinArms/DEGlist_matches_upinarms_summer20227controlV7armdrop.tab", sep = "\t", row.names = T, quote = FALSE, col.names = TRUE)
```

Write out 01/01/2024. 