---
title: "13-compareUpinArms-2022"
output: html_document
date: "2024-01-02"
---
Rmd to compare DEGlist from Up in Arms 2015 to Summer 2022 DEG list comparing 7control and 7 exposed arm drop stars. 

Load packages needed:
```{r}
library(tidyverse)
library(dplyr)
```

Read in Up in Arms DEGlist from https://github.com/sr320/eimd-sswd/blob/master/precompiled_wd/Phel_DEG_uniprot.tab. I downloaded the file and put in the data directory for this repository. 

```{r}
uiadeg <- read.delim("../data/Phel_DEG_uniprot_UpinArms.tab", sep = '\t', header = TRUE)
head(uiadeg)
```

Pull out just the first two columns:
```{r}
uiatuni <- select(uiadeg, "transcript_ID", "uniprotID")
head(uiatuni)
```

Read in the DEGlist with uniprot from the summer 2022 deseq2 comparison of 7controls vs 7 exposed arm drop stars:
```{r}
cvad <- read.delim("../analyses/08-deseq2/DEGlist_control_v_armdrop_7x7_genecounts_blast.tab")
head(cvad)
```

`join` this list with the uiatuni list based on the column "uniprotID"
```{r}
cvaduia <- left_join(cvad, uiatuni, by = "uniprotID")
head(cvaduia)
```

Remove rows with no input in uniprotID column:
newdf <- df[!is.na(df$name),]
```{r}
cvaduiashort <- cvaduia[!is.na(cvaduia$uniprotID),]
head(cvaduiashort)
```

Remove rows that don't have a Up in Arms transcript ID:
```{r}
cvaduiashorter <- cvaduiashort[!is.na(cvaduiashort$transcript_ID),]
head(cvaduiashorter)
```

It is now 1,727 rows long. 

Read this file into a analyses/13-compare-UpinArms:
```{r}
#write.table(cvaduiashorter, "../analyses/13-compare-UpinArms/DEGlist_matches_upinarms_summer20227controlV7armdrop.tab", sep = "\t", row.names = T, quote = FALSE, col.names = TRUE)
```

Write out 01/01/2024. 


## Combine this table with the uniprot-SP-GO table to get Go IDs:

read in table: 
```{r}
combodeglist <- read.delim("../analyses/13-compare-UpinArms/uniqueDEGs_overlap-UpinArms_Summ22_7controlV7armdrop.txt")
head(combodeglist)
```

read in uniprot-SP-GO:
Note: I copied it into code directory from another directory I had it in. 

```{R}
uniprotSPGO <- read.delim("../code/uniprot-SP-GO.sorted", header = FALSE)
head(uniprotSPGO)
```

#joining blast with uniprot annoation file and reducing to three columns UniprotID, Query, All GO terms
```{bash}
join -t $'\t' \
combodeglist \
uniprot-SP-GO.sorted \
| cut -f1,2,14 \
> combodeglist-annot.tab
```
# `join` `../analyses/14-annotate-deglist-blastGO/DEGlist_7controlV7armdrop_blastGOslim_count_uniprot.tab` with `../analyses/13-compare-UpinArms/uniqueDEGs_overlap_UpinArms_Summ22_7controlV7armdrop.txt`

```{r}
deglistblastuniprot <- read.delim("../analyses/14-annotate-deglist-blastGO/DEGlist_7controlV7armdrop_blastGOslim_count_uniprot.tab", sep = '\t')
head(deglistblastuniprot)
```

```{r}
uniqDEGsUp <- read.delim("../analyses/13-compare-UpinArms/uniqueDEGs_overlap-UpinArms_Summ22_7controlV7armdrop.txt", sep = '\t')
head(uniqDEGsUp)
```

`join` by "geneID"
```{r}
deglistblastuniprotUp <- left_join(deglistblastuniprot, uniqDEGsUp, by = "geneID")
head(deglistblastuniprotUp)
```

write out
```{r}
#write.table(deglistblastuniprotUp, "../analyses/13-compare-UpinArms/DEGlist_matches_upinarms_summer20227controlV7armdrop_DEGlistBLastuniprot.tab", sep = "\t", row.names = T, quote = FALSE, col.names = TRUE)
```
wrote out 2024-01-02



