---
title: "14-annotate-deglist-blastGO"
output: html_document
date: "2024-01-02"
---
Rmd to annotate DEGlist from SICB analyses with the BLAST and GO. 

```{r}
deglist <- read.delim("../analyses/08-deseq2/DEGlist_control_v_armdrop_7x7.tab", sep = '\t')
head(deglist)
```

Make rownames into column called "geneID"
```{r}
library(tibble)
deglist <- tibble::rownames_to_column(deglist, "geneID")
head(deglist)
```


```{r}
blastGOslim <- read.delim("../analyses/14-annotate-deglist-blastGO/Blastquery-GOslim.tab", sep = '\t', header = FALSE)
head(blastGOslim)
```
Set column names:
```{r}
blastGOslim <- setNames(blastGOslim, c("geneID", "GOID", "Biological_Process", "P"))
head(blastGOslim)
```

`join` the two files based on column "geneID"
```{r}
deglistblastGOslim <- left_join(deglist, blastGOslim, by = "geneID")
head(deglistblastGOslim)
```

write out file:
```{r}
#write.table(deglistblastGOslim, "../analyses/14-annotate-deglist-blastGO/DEGlist_7controlV7armdrop_blastGOslim.tab", sep = '\t', row.names = FALSE, quote = FALSE)
```

wrote out 2024-01-02



`join` the above list with the DEGlist of this comparison that has the BLAST output and the gene counts:
```{r}
degblastcount <- read.delim("../analyses/08-deseq2/DEGlist_control_v_armdrop_7x7_genecounts_blast.tab", sep = '\t')
head(degblastcount)
```

`join` by "geneID":
```{r}
degblastcountGOslim <- left_join(degblastcount, deglistblastGOslim, by = "geneID")
head(degblastcountGOslim)
```

write out:

```{r}
#write.table(degblastcountGOslim, "../analyses/14-annotate-deglist-blastGO/DEGlist_7controlV7armdrop_blastGOslim_count.tab", sep = '\t', row.names = FALSE, quote = FALSE)
```

wrote out 2024-01-02

annotate with uniprot-SP-GO:
```{r}
uniprot <- read.delim("../analyses/14-annotate-deglist-blastGO/uniprot-SP-GO.sorted", header = FALSE)
head(uniprot)
```
 rename first column as uniprotID:
```{r}
uniprot <- setNames(uniprot, c("uniprotID", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9", "V10", "V11", "V12"))
head(uniprot)
``` 
left join with degblastcountGOslim
```{r}
degblastcountGOslimuniprot <- left_join(degblastcountGOslim, uniprot, by = "uniprotID")
head(degblastcountGOslimuniprot)
```


```{r}
#write.table(degblastcountGOslimuniprot, "../analyses/14-annotate-deglist-blastGO/DEGlist_7controlV7armdrop_blastGOslim_count_uniprot.tab", sep = '\t', row.names = FALSE, quote = FALSE)
```

wrote out 2024-01-02


